some title

### 2018-01-31 2018/images/f_dist_def.tex
geometry of r.v.s

\begin{tikzpicture}[scale=5,tdplot_main_coords]

\coordinate (O) at (0,0,0);
\coordinate (xc) at (1,0);
\coordinate (yc) at (0,2);
\coordinate (un) at (0,0,4);

\tdplotsetcoord{S}{2}{30}{0}  % икс
\tdplotsetcoord{P}{4}{30}{90} % игрек
\tdplotsetcoord{Z}{3.4641}{0}{0} % 3.4641 = \sqrt{4^2 - 2^2}

%\draw[color=white, fill=gray!10] (xc) to[out=90,in=90] (S)  to[out=90,in=180]  (un) -- (O) -- cycle;
%\draw[color=white, fill=blue!10] (yc) to[out=90,in=90] (P) to[out=90,in=0] (un) -- (O) -- cycle;

\draw[color=white, fill=gray!5] (S)  to[out=90,in=180]  (un) -- (O) -- cycle;
\draw[color=white, fill=blue!5] (P) to[out=90,in=0] (un) -- (O) -- cycle;

%\draw[color=white, fill=gray!10, pattern=dots] (S)  to[out=90,in=180]  (un) -- (O) -- cycle;
%\draw[color=white, fill=blue!10, pattern=north east lines] (P) to[out=90,in=0] (un) -- (O) -- cycle;

\draw[thick,->] (O) -- (xc) node[anchor=north east]{$x - \bar{x} \cdot \mathbf{1}$};
\draw[thick,->] (O) -- (yc) node[anchor=north west]{$y - \bar{y} \cdot \mathbf{1}$};
\draw[thick,->] (O) -- (un) node[anchor=south]{$\mathbf{1}$};

\tdplotdrawarc{(O)}{0.1}{0}{90}{anchor=north}{$\phi$}

\draw[-stealth,thick] (O) -- (S) node[anchor=north east]{$x$};
\draw[-stealth,thick] (O) -- (P) node[anchor=north west]{$y$};

\draw[dashed] (S) -- (Sx);
\draw[dashed] (P) -- (Py);

\tkzMarkRightAngle[size=0.05](P,yc,O);
\tkzMarkRightAngle[size=0.05](S,xc,O);

\end{tikzpicture}

\end{document}

>>>>

\begin{tikzpicture}[tdplot_main_coords]

\coordinate (O) at (0,0,0);
\coordinate (un) at (0,0,4);
\coordinate (nun) at (0,0,-2.2);
\coordinate (hun) at (0,0,3.2);
\coordinate (z) at (0,3,5);
\coordinate (hz) at (-1.05,2.05,0);

\draw[->] (O) -- (un) node[anchor=east]{$L_1$};
\draw[dashed] (O) -- (nun);
\draw (0,0) ellipse (4.5cm and 1.5cm) node[anchor=east, left=3.2cm]{$L_2$};

\draw[->] (O) -- (z) node[anchor=south]{$z$};
\draw[->] (O) -- (hz) node[anchor=north]{$H_2 z$};
\draw[->] (O) -- (hun) node[anchor=east]{$H_1 z$};
\draw[dashed] (hz) -- (z);
\draw[dashed] (z) -- (hun);
\tkzMarkRightAngle[size=0.15](z,hz,O);
\tkzMarkRightAngle[size=0.15](O,hun,z);

\end{tikzpicture}
\end{document}

### 2018-01-31 2018/paper.tex
geometry of r.v.s

@@

\subsection{Sample correlation}

@@

The most crucial part in defining correlation geometrically is definig the dot product as it enables to compute the length of a vecotr:
\[
|\vec{a}| = \sqrt{\langle \vec{a},  \vec{a}\rangle}
\]
and the angle between any two vectors:
\[
\cos(\vec{a}, \vec{b}) = \frac{\langle \vec{a},  \vec{a}\rangle}{|\vec{a}| |\vec{b}|}
\]
Now we define scalar product of two vectors
$x =
\begin{pmatrix}
x_1 \\
\vdots \\
x_n
\end{pmatrix}$
and
$y =
\begin{pmatrix}
y_1 \\
\vdots \\
y_n
\end{pmatrix}$
as a sample covariation between them:
\[
\langle x, y \rangle = \sCov(x, y)
\]
The main characteristics of a vector are its length and direction.
So, we introduce the length
\[
\sqrt{\sCov(x,x)} = \sqrt{\sVar(x)} = \sigma_x
\]
and the angle between two random vectors
\[
\cos(x,y) = \frac{\sCov(x,y)}{\sqrt{\sVar(x)\sVar(y)}} = \sCorr(x,y)
\]
Note that from the definition of the angle it follows that the sample correlation can range from $-1$ to $1$.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.4\linewidth]{images/corr_def.pdf}
\caption{Vector $x$ of length $\sigma_x$ and random vector $y$ of length $\sigma_y$, $\cos \varphi$ is the angle between $x$ and $y$.}
\label{fig:corr_def}
\end{center}
\end{figure}

Another important geometrical tool is projection. Recall that for any two vectors the scalar product $\langle \vec{a}, \vec{b} \rangle$ can be interpreted as the length of projected $\vec{b}$ multuplied by the length of $\vec{a}$.
The projection itself is $cos(\vec{a}, \vec{b}) \vec{b}$.
Same holds for the vectors sampled from some distribution.
The projection of such a vector $y$ onto $\{cx| c \in \mathbb{R}\}$ is $\hat y = \sCorr(x,y) \cdot y$.

@@

Looking at Figure~\ref{fig:corr_proj}, we can interpret the square of sample correlation coefficient.
Using the fact that $cos^2 \varphi$ is the squared ratio of the leg adjacent to $\varphi$ to hypotenuse, we can conclude that
\[
\sCorr^2(x,y) = \frac{\sVar(\hat y)}{\sVar(y)}
\]
as the variance of a vector is associated with the square of its length.
Thus, the sample correlation coefficient squared shows the fraction of variance in $y$ which can be explained
with the most similar vector proportional to $x$.

@@

>>>>

\section{Geometric interpretation of random variables}

The fundamental part to start with is defining the geometric properties of random
variables using some concepts of linear algebra.

Consider the vector space which consists of all random variables with finite
mean and variance.
We will regard each point in this space (or vector that correponds to that point
in terms of linear algebra) a random variable.
We define the scalar product of two random variables $X$ and $Y$ to be
\[
\langle X, Y \rangle = \Cov(X,Y).
\]
It is of no difficulty to check that the definition satisfies the properties
of scalar product assuming that $X$ and $Y$ are the same random variables
if $\P(X=Y)=1$.

Having defined the inner product, we are now able to introduce the squared
length of a random variable $X$ which is
\[
\lVert X \rVert^2 = \langle X, X \rangle = \Cov(X,X) = \Var(X),
\]
so the length is simply the square root of this expression, i.e., the~standard
deviation of $X$ ($\sigma_X$).

Recall that for any non-random vectors $\vec{a}$ and $\vec{b}$ the angle
between them is calculated with the formula
\[
\cos(\vec{a}, \vec{b}) = \frac{\langle \vec{a},  \vec{a}\rangle}{|\vec{a}| |\vec{b}|}.
\]
The same applies for the random variables and it is already clear that
two random variables are uncorrelated if and only if their scalar product
equals to $0$. Additionally, it means that these two random variables
are orthogonal in the vector space.

The analogue for $cos(\vec{a}, \vec{b})$ in the vector space of all the random
variables is the correlation between two of them:
\[
\Corr(X,Y) = \frac{\Cov(X,Y)}{\sqrt{\Var(X)\Var(Y)}} = \frac{\langle X, Y \rangle}{\sqrt{\lVert X \rVert^2 \lVert Y \rVert^2}}.
\]
From the equivalence of $\Corr(X,Y)$ to the $\cos(\vec{a}, \vec{b})$ it
automatically follows that the correlation coefficient can range from $-1$ to $1$.

A useful property of the geoemtry of random variables is that all the
geometric theorems still hold. For instance, the Pythogorean theorem can
be formulated as follows: if the ranadom variables $X$ and $Y$ are uncorrelated
(which implies that they are orthogonal), then the variance of their sum equals
the sum of their variances:
\[
\Var(X + Y) = \sigma^2_{X+Y} = \sigma^2_{X} + \sigma^2_{Y} = \Var(X) + \Var(Y).
\]
Translated to the non-random language, assumption of uncorrelatedness correspnds
to the right triangle setting, the variance of the sum of two random variables
stands for the hypotenuse squared and the sum of the variances is the sum of
the legs squared.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.4\linewidth]{images/rv_pythagorean.pdf}
\caption{The Pythagorean theorem for random variables $X$ and $Y$.}
\label{fig:rv_pyth}
\end{center}
\end{figure}

Another important geometric tool is projection.
Recall that for any two vectors the scalar product $\langle \vec{a}, \vec{b} \rangle$
can be interpreted as the length of projected $\vec{b}$ multuplied by the length of $\vec{a}$.
The projection itself is $cos(\vec{a}, \vec{b}) \vec{b}$.
Same holds for the random variables.
The projection of such a random variable $Y$ onto $\{cX| c \in \mathbb{R}\}$ is
$\hat Y = \Corr(X,Y) \cdot Y$.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.4\linewidth]{images/rv_proj.pdf}
\caption{The projection of a random variable $Y$ onto the line spanned by
a random varibale $X$.}
\label{fig:rv_proj}
\end{center}
\end{figure}

Note that the squared lengths of the leg adjacent to $\varphi$ and the
hypotenuse are $\Var(\hat Y)$ and $\Var(Y)$.
So, the Figure~\ref{fig:rv_proj} gives a useful expression for the correlation
coefficient squared:
\[
\Corr^2(X,Y) = \frac{\Var(\hat Y)}{\Var(Y)}.
\]

\subsection{The law of iterated expectations}

\marginnote{
Here is the proof for the case when $X$ and $Y$ are both discrete. Let $\E(Y|X) = g(X)$.
\begin{align*}
\E(g(X)) &= \sum_x g(x) \P(X=x) \\
&= \sum_x \left( \sum_y y \P(Y=y|X=x) \right) \P(X=x) \\
&= \sum_x \sum_y y  \P(X=x) \P(Y=y|X=x)  \\
&= \sum_y y \sum_x \P(X=x, Y=y) \\
&= \sum_y y \P(Y=y) \\
&= \E(Y)
\end{align*}
The proof in case of continous random variables is absolutely analogous.
}

\begin{theorem}
For any random variable $X$ and $Y$,
\[
\E(\E(X|Y)) = \E(Y).
\]
\end{theorem}

\begin{proof}

Consider the vector space of all the random variables.  The random variable
which can be described as functions $h(X)$ of $X$ form a subspace of that vector
space, represented as a plane $\alpha$ in Figure~\ref{fig:adams}.
Another subspace is a subspace of constants, denoted as a vector $\mathbf{1} \in \alpha$.

In order to obtain $\E(Y|X)$, first, we need to project $Y$ onto the subspace
corresponding to $X$. As a result of this step, we get $\E(Y|X)$ — the function
of $X$ that predicts $Y$ the best. Next, projecting $\E(Y|X)$ onto the space
of all constants, we end up with $E(Y)$.

Notice that the vector $Y - \E(Y|X)$ (which is also called the residual) is
perpendicular to the plane $\alpha$. Moreover, the vecotr $\E(Y|X) - E(Y)$ is
perpendicular to the vector of constants $\mathbf{1}$. Thus, we can apply
the theorem of three perpendiculars and conclude that the vector $Y - \E(Y)$ is
also perpendicular to the vector of constants $\mathbf{1}$.

So, we showed that the expectation of the random variable $Y$ can be obtained either
in two steps or by its direct projection onto the subpace of constans.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\linewidth]{images/adams.pdf}
\caption{The law of iterated expectations. Equivalence of the two-step projecttion
and direct projection of $Y$ onto $\mathbf{1}$.}
\label{fig:adams}
\end{center}
\end{figure}

\subsection{MSE decomposition}

\marginnote{
\begin{align*}
MSE(\hat \theta) &= \E((\hat \theta - \theta)^2) \\
&= \E \left[ \left( \hat \theta - \E(\hat \theta) +  \E(\hat \theta) - \theta \right) \right] \\
&= \E \left[ \left( \hat \theta - \E(\hat \theta) \right)^2 + 2 \left( (\hat \theta - \E(\hat \theta)) (\E(\hat \theta) - \theta ) \right) \right. \\
&+ \left. \left( \E(\hat \theta) - \theta  \right)^2 \right] \\
&= \E \left[\left( \hat \theta - \E(\hat \theta) \right)^2 \right] \\
&+ 2 \E \left[\left( (\hat \theta - \E(\hat \theta)) (\E(\hat \theta) - \theta ) \right) \right] \\
&+ \E \left[\left( \E(\hat \theta) - \theta  \right)^2 \right] \\
&= \E \left[\left( \hat \theta - \E(\hat \theta) \right)^2 \right] \\
&+ 2 (\E(\hat \theta - \E(\hat \theta))) \E(\hat \theta - \E(\hat \theta)) \\
&+ \E \left[\left( \E(\hat \theta) - \theta  \right)^2 \right] \\
&= \E \left[\left( \hat \theta - \E(\hat \theta) \right)^2 \right] + \E \left[\left( \E(\hat \theta) - \theta  \right)^2 \right] \\
&= \Var(\hat \theta) + \E \left[\left( \E(\hat \theta) - \theta  \right)^2 \right]
\end{align*}
}

\begin{theorem}
The mean squared error of an estimator $\hat \theta$ with respect to an unknown
parameter $\theta$ defined as $MSE(\hat \theta) = \E((\hat \theta - \theta)^2)$
can be decomposed into the sum of the variance of the estimator and its squared bias:
\[
MSE(\hat \theta) = \Var(\hat \theta) + \E \left[\left( \E(\hat \theta) - \theta  \right)^2 \right]
\]
\end{theorem}

\begin{proof}
We start with a random variable $\theta$ and its estimate $\hat \theta$ in the
vector space. We know that an unbiased estimator's projection would be exactly
the vector representing $\theta$. However, in general it does not have to and
Figure~JJJJ illustrates this case: the projection of the estimator falls onto
the line spanned by the vector $\theta$.

Connecting vectors $\theta$ and $\hat \theta$, we obtain the right triangle which
legs are $\hat \theta - \E(\hat\theta)$, $\E(\hat\theta) - \theta$ and the
hypotenuse $\hat \theta - \theta$.
Applying the Pythagorean theorem, we finish the proof:
\begin{align*}
\lVert \hat \theta - \theta \rVert^2 &= \lVert \hat \theta - \E(\hat \theta) \rVert^2  + \lVert \E(\hat \theta) - \theta \rVert^2 \\
\E((\hat \theta - \theta)^2) &= \E((\hat \theta - \E(\hat \theta))^2) + \E((\E(\hat \theta) - \theta)^2) \\
MSE(\hat \theta) &= \Var(\hat \theta) + \E((\E(\hat \theta) - \theta)^2)
\end{align*}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\linewidth]{images/mse_decomposed.pdf}
\caption{Decomposition of mean squred error into the variance and the bias squared.}
\label{fig:mse_decomposed}
\end{center}
\end{figure}

@@

The concepts discussed in the following section could also be presented in
random variables instead of sample ones. As the geometry of sample variables
is almost of no difference comparing to the random ones, the logic of
all the theorems is also the same.

\subsection{Geometry of sample variables}

@@

In the same manner which was performed in Section 1, we define
the scalar product of two sample variables
$x =
\begin{pmatrix}
x_1 \\
\vdots \\
x_n
\end{pmatrix}$
and
$y =
\begin{pmatrix}
y_1 \\
\vdots \\
y_n
\end{pmatrix}$
as a sample covariation between them:
\[
\langle x, y \rangle = \sCov(x, y).
\]
The main characteristics of a vector are its length and direction.
Again, we introduce the length
\[
\sqrt{\sCov(x,x)} = \sqrt{\sVar(x)} = \sigma_x
\]
and the angle between two sample variables
\[
\cos(x,y) = \frac{\sCov(x,y)}{\sqrt{\sVar(x)\sVar(y)}} = \sCorr(x,y).
\]
Note that from the definition of the angle it follows that the sample correlation can range from $-1$ to $1$.

Completely analogus to the case of random variables,
the projection of such a sample variable $y$ onto $\{cx| c \in \mathbb{R}\}$ is $\hat y = \sCorr(x,y) \cdot y$.

@@

Looking at Figure~\ref{fig:corr_proj}, we can interpret the square of sample correlation coefficient.
Using the fact that $cos^2 \varphi$ is the squared ratio of the leg adjacent to $\varphi$ to hypotenuse, we can conclude that
\[
\sCorr^2(x,y) = \frac{\sVar(\hat y)}{\sVar(y)},
\]
as the variance of a vector is associated with the square of its length.
Thus, the sample correlation coefficient squared shows the fraction of variance in $y$ which can be explained
with the most similar vector proportional to $x$.

@@

Consider a random vector $z \in \mathbb{R}^n$ which components are independent
and follow standard normal distribution, $z_i \sim \mathcal{N}(0,1)$.
Consider also a fixed $k$-dimensional subspace in $\mathbb{R}^n$ $L$.
Let the projection of vector $z$ onto the subspace $L$ be $\hat z$ and
its length squared $Q$
\[
Q = \lVert \hat z \rVert^2 = \langle \hat z, \hat z \rangle = \hat z^T \hat z
\]
Then $Q$ follows the chi-squared distribution with $k$ degrees of freedom.
